---
title: "Unit testing in R"
author: "Davide Risso"
date: "5/11/2017"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)
```

## Outline

- What is unit testing?
- Why is unit testing important?
- The `testthat` package
- Test coverage and the `covr` package
- Unit tests and Continuous Integration
- Travis CI and R Travis

## Useful resources

- Hadley Wickham's [R packages book](http://r-pkgs.had.co.nz)
       - In particular the "Testing" chapter
- ["Writing R extensions" manual](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
- [R Travis](https://docs.travis-ci.com/user/languages/r/)
- [The `covr` vignette](https://cran.r-project.org/web/packages/covr/vignettes/how_it_works.html)

## What is unit testing?

From wikipedia...

In computer programming, unit testing is a software testing method by which __individual units of source code__ [...] are tested to determine whether they are fit for use.

In R packages 

`units of codes == functions`

## Why is unit testing important {.build}

- __When creating a package__, to make sure that the implementation of the method is correct.

Hence, test the code on some cases for which the answer is known and make sure that it can be retrieved.

- __When modifying the code__ to introduce new features or to fix a bug, to make sure that the fix doesn't break the implementation.

Hence, running the tests every time we release the code.

__Ideally, at every commit!__

## Why should I bother?

The reality is that you already test the code, just not in a __formal__, __systematic__ way.

The problem with informal testing is that you have to do it from scratch every time. It's very easy to forget some tests and break existing code.

## Advantages of formal testing

(According to Hadley Wickham)

> 1. Fewer bugs. If you write a test for every bug that you fix, old bugs will not come back in the future.
> 2. Better code structure. Writing tests forces you to break up complicated code into separate functions.
> 3. Easier restarts. Create a failing test at the end of the session and running the test will tell you what to do next.
> 4. Robust code. If all the major part of the package are tested, you can confidently change many functionality without the fear of breaking the package.

## Unit testing in R

Testing is really just writing some code in a script and running it every time you make a major change to the package.

One of the best features of R is `R CMD check` and luckily there is a way to add your tests to your `R CMD check` routine.

All it takes is to add your script(s) to the `tests/` directory.

[Example from my RUVSeq package](https://github.com/drisso/RUVSeq/tree/master/tests).

## The `testthat` package

As with most of the things you want to do in R, Hadley Wickham has a package for that...

It provides a set of utility functions such as `expect_equal()`, `expect_error()`, etc. that make testing easier.

## The `testthat` package

To set up your package to use testthat, run:

```{r eval=FALSE}
devtools::use_testthat()
```

This will:

1. Create a tests/testthat directory.
2. Add testthat to the Suggests field in the DESCRIPTION.
3. Create a file tests/testthat.R that runs all your tests when `R CMD check` runs.

You can then add all your tests as scripts in the `tests/testthat/` folder.

[Example from my zinbwave package](https://github.com/drisso/zinbwave/tree/master/tests).

## Live demo

1. Create a simple R package.
2. Add some tests.
3. Run `R CMD check`.

## Test coverage and the `covr` package

I find it extremely useful to know how much of my code is covered by at least one unit test.

The `covr` package by Jim Hester does just that: it runs through the functions in a package and checks whether they are covered by your tests.

This is as simple as

```{r eval=FALSE}
covr::package_coverage()
```

Websites like `codecov.io` and `coveralls.io` provide a nice graphical interface.

[Example](https://codecov.io/gh/drisso/zinbwave).

## Unit testing and Continuous Integration

`R CMD check` automatically checks your code for common problems.

It is mandatory to pass `R CMD check` without warnings before submitting to CRAN or Bioconductor.

Even for your personal package, it is good practice to check your package often.

How often? As often as you can... i.e., after every commit!

## Travis C.I.

All it takes to use Travis C.I. (after setting up an account) is to write a few instructions in a yaml file named `.travis.yml` to be added to your GitHub repository.

[Example](https://github.com/drisso/zinbwave/blob/master/.travis.yml).

Travis is free for open source projects (up to 50 minutes for every push).

## Linking test coverage to your Travis C.I. build

The `covr` package has a `codecov()` and a `coveralls()` function that can be run on Travis to generate a report to be uploaded to `codecov.io` or `coveralls.io` respectively.

All it takes is to add the following lines to the Travis yaml file.

```{r eval=FALSE}
after_success:
  - Rscript -e 'library(covr); codecov()'
```

## Live demo

1. Create a GitHub repository for the R package.
2. Add a .travis.yml file to the repositort.
3. Push changes to trigger Travis build.
4. Add repo to codecov.io to explore test coverage.
5. Badges! :)

